Shapes analysis for time series.

### 2024-05-21 NEURIPS24/paper.tex
update remarque laurent

%Idée de parler d'inter individualité, spécialiser davantage
%k shape, shapelets , ye2009time
%\thi{à reformuler pour faire moins stupide}

%\sam{Aller droit au but, se servir de la section related works pour détailler, notamment enlever le premier paragraphe dire directement que l'on veut faire du unsupervised represnetation learning for time series mais en s'intéressant par ticulièrement à la notion de shape.
%Parler du contexte biomédical, définir la shape. Parler de shape analysis, introduire notre problème de représentation, dire que nous on propose une structure adapté aux time series.}

% In this paper, we propose an unsupervised representation learning method (URL) for time series \cite{trirat2024universal,meng2023unsupervised} by giving a special attention to their \textit{shapes} for studying inter-variability in biomedical applications.
% In particular, the method handle time series with \textbf{irregular sampling} and \textbf{different sizes}.
% This work belongs to the shape analysis literrature: %which focus on the statistical analysis of various mathematical objects invariant under rotations, dilations, or time parameterization.
%  we extend Large Deformation Diffeomorphic Metric Mapping (LDDMM) \cite{beg2005computing,vaillant2004statistics} to the case of time series data by applying difformations to their graph.
% However, time series graph have more structure than 1D curves:
%  a graph time series transformation by a general diffeomorphism is not always a graph time series.
% This work is at the crossroad of URL for time series and shape analysis; we hope it will benefit to both communities.
%A time series is a set of values indexed by time \cite{trirat2024universal}.
%This type of data is prevalent in biology and medicine for recording physiological time series , as well as in the broader context of the Internet of Things sensors \cite{fathy2018large} and measurements in physical systems \cite{luo2021deep}.
 %However, due to the nature of such data, which have different sample sizes and incorporate a temporal evolution, classical statistical analysis, and machine learning methods need refinement \cite{anderson2011statistical}.
 %One option to tackle this issue is to derive a feature representation of time series, which depends on the problem at hand \cite{trirat2024universal}.

 Our goal is to analyze the inter-individual variability within a time series dataset, an approach of prime interest in physiological contexts \cite{guscelli2019importance,wang2016research,bar2012studying,germain2023unsupervised}.
  More specifically, we aim to find an unsupervised features representation method that encodes individual time series specificities compared to a reference one.
In physiology, studying the different \textit{shapes} in a time series related to biological phenomena and their variations according to individual or pathology is common.
However, a \textit{shape} has no clear definition; it is more an intuitive way to speak about the silhouette of a pattern in a time series. In this paper, we refer to as the shape of a time series, the graph of this signal.

Although a community structure with representatives can be learned in an unsupervised way \cite{trirat2024universal,meng2023unsupervised} using contrastive loss \cite{franceschi2019unsupervised,tonekaboni2021unsupervised,meng2023unsupervised} or similarity measures \cite{asgari2023clustering,germain2023unsupervised,paparrizos2015k,ye2009time},
studying the inter-individual variability of shapes within a cluster \cite{niennattrakul2007inaccuracies,shirato2023identifying} is still an open problem in URL and even more for \textit{irregularly sampled} time series with \textit{variable lengths}.
     %In this paper, we aim to go beyond group structure by giving an unsupervised data-driven method that represents the specificity of each individual's shape \cite{vidal2021perceptual}.
  %For instance, the shape of a heartbeat in an Electrocardiogram (ECG) -the pattern- has been well studied \cite{madona2021pqrst} with the PQRST decomposition (see figure\sam{fill}).
  %There exist several methods to detect and segment patterns in a time series (\sam{mettre ref thibuat} ).
   %The goal is then to study their shape according to the property of the individual.

  %  In most clinical studies, depending on expert knowledge, people inspect hand-crafted features such as the mean, wavelets decomposition, or temporal correlation \cite{anderson2011statistical} to keep interpretability.
  %  However, this approach is time-consuming and potentially biased by priors, especially in exploratory research.
  %  Given the success of neural networks \cite{bishop1994neural}, features can now be automatically extracted, depending only on the data, by choosing an objective to optimize and a neural architecture tailored to the problem \cite{ismail2019deep,blazquez2021review,torres2021deep}, even in an unsupervised way \cite{meng2023unsupervised}.
  %  Moreover, depending on the method, it is still possible to have an interpretability of the features.
  %   For instance, among individuals, a contrastive clustering method can be applied to spot community structures \cite{franceschi2019unsupervised,tonekaboni2021unsupervised,meng2023unsupervised}.
  %   More generally, interpretability \cite{yuan2019wave2vec,zhang2018visual,ismail2020benchmarking} is possible by visualizing the shapes of representative patterns to which the model is sensitive.
  %   Other methods, based on similarity measures \cite{asgari2023clustering,germain2023unsupervised,paparrizos2015k}, also propose to perform clustering on shapes in time series or to find some representatives \cite{ye2009time}.

    %Although a community structure with representatives can be learned in an unsupervised way,
    % studying the inter-individual variability of shapes within a cluster \cite{niennattrakul2007inaccuracies,shirato2023identifying} is still an open problem.
     %In this paper, we aim to go beyond group structure by giving an unsupervised data-driven method that represents the specificity of each individual's shape \cite{vidal2021perceptual}.% \thi{In an individual cluster ?}

% Alternatively, one can represent time series with an embedding that can be processed using conventional tools. % or we can represent the time series with an embedding which can be treated by classical tools.

% Depending on the tasks, e.g classification \cite{ismail2019deep}, anomaly detection \cite{blazquez2021review}, forecasting \cite{torres2021deep} or clustering , a meaningful vector representation of time series can be derived.
%  Initially, in this research, handcrafted features were computed from time series, such as the mean, wavelets \cite{percival2000wavelet}, or Fourier coefficients \cite{bloomfield2004fourier}, depending on expert knowledge.
%   However, this approach is time-consuming and potentially biased by priors.
%   Observing the success of neural networks, features are automatically extracted by choosing an objective to optimize and a neural architecture tailored to the problem.
%    Taking a set of time series as input, features can be identified by solving specific supervised tasks—classification \cite{ismail2019deep}, anomaly detection \cite{blazquez2021review}, forecasting \cite{torres2021deep}— or in an unsupervised manner by deriving labels from the inputs \cite{meng2023unsupervised}.
%    \thi{Rajouter une image, et une phrase par rapport au temps}
%    Although the deep learning approach is particularly effective for supervised tasks, annotated datasets are rare, making unsupervised methods more practical.
%    These methods have proven effective in identifying different patterns in the data and providing a vector representation that segregates well into clusters \cite{meng2023unsupervised}.
%     However, they may lack precision on small and homogeneous datasets with only one typical shape, often arising from clinical experiments addressing a specific phenomenon (refer to figures).
%      While current unsupervised methods separate representations into clusters, our focus is on the inter-individual variability within one cluster.
%       Typically, we aim to perform Principal Components Analysis (PCA) \cite{bro2014principal} on individual patterns with similar shapes. The representation is dependant on the input dataset.

%This type of question has already been addressed in shape analysis \cite{vaillant2004statistics}, a community focusing on the statistical analysis of various mathematical objects invariant under rotations, dilations, or time parameterization.
%However, time series data with their specific structure has not been properly adressed.
Our work focuses explicitly on learning shape-based representation of time series. First, we propose not to see the shape of a time series through its curve $\{s_t:\eqsp t\in\msi\}$, but rather through its graph $\msg(s)=\{(t,s(t)):\eqsp t\in \msi\}$.
   Then, building on the shape analysis literature \cite{beg2005computing,vaillant2004statistics}, we follow the Large Deformation Diffeomorphic Metric Mapping (LDDMM) framework \cite{beg2005computing,vaillant2004statistics} to analyze these graphs. The idea is to represent each element $\msg(s^j)$ of a dataset $(s^j)_{j\in[N]}$ as the transformation of a reference graph $\msg(\mathbf{s}_0)$ by a diffeomorphism $\phi_j$, i.e. $\msg(s^j) \sim \phi_j . \msg(\mathbf{s}_0)$ .
    The diffeomorphism $\phi_j$ is learned by integrating an ordinary differential equation parameterized by a Reproducing Kernel Hilbert Space (RKHS).
     The parameters $(\alpha_j)_{j\in[N]}$ encoding the diffemorphisms $(\phi_j)_{j\in[N]}$ yield the representation features of the graphs $(\msg(s^j))_{j\in[N]}$. Finally, these shape-encoding features can feed any statistical or machine-learning model.
%  First, we propose not to see time series through their curve $\{s_t:\eqsp t\in\msi\}$, but through their graph $\msg(s)=\{(t,s(t)):\eqsp t\in \msi\} $. 
%  Then, building on the shape analysis litterature \cite{beg2005computing,vaillant2004statistics}, we follow the Large Deformation Diffeomorphic Metric Mapping (LDDMM) framework \cite{beg2005computing,vaillant2004statistics} to analyze these graphs.
%   The idea is to represent each element $(\msg(s^j))_{j\in[N]}$ of the dataset as the transformation of a reference graph $\msg(\mathbf{s}_0)$ by a diffeomorphism.
%   Then, the diffeomorphism is learned by integrating an ordinary differential equation parameterized by a Reproducing Kernel Hilbert Space (RKHS).
%    The parameters $(\alpha_j)_{j\in[N]}$ encoding the diffemorphisms $(\phi_j)_{j\in[N]}$ yield the representation features of the graphs $(\msg(s^j))_{j\in[N]}$.
%     Finally, these features encoding the shapes can feed any statistical or machine learning model as in URL.

@@

     LDDMM framework is a relevant shape analysis tool to represent curves as depicted in \cite{glaunes2008large}. However, graphs of time series are a well-structured type of curve due to the inclusion of the temporal dimension that requires specific care (\Cref{fig:diffeo}).
      Closely related, \cite{qiu2009time} tracks anatomical shape changes in serial images using LDDMM but distinguishes from us by including the temporal evolution at a higher level: the goal is to perform longitudinal data modeling.

       Leaving the LDDMM representation, \cite{srivastava2010shape,heo2024logistic} addresses the representation of curves with the Square-Root Velocity (SRV) representation.
        However, the SRV representation is applied after a reparametrization of the temporal dimension on the unit length segment.
         Consequently, the graph structure of the time series is not respected, and the original time evolution of the time series is not encoded in the final representation.
          Very recently, in a functional data analysis framework, a paper \cite{wu2024shape} (Shape-FPCA) improved by representing the original time evolution.
           Nevertheless, this method is made for \textit{continuous objects} and only applies to time series of \textit{same length}, making the estimation more sensitive to noise.
  %However, the case of time series has not been adequately addressed, considering the special attention the time axis requires.
  %For instance, the SRV representation applies after a reparametrization in time, such that the original time evolution of the time series is not represented in the final features.
% Depending on the tasks, e.g classification \cite{ismail2019deep}, anomaly detection \cite{blazquez2021review}, forecasting \cite{torres2021deep} or clustering , a meaningful vector representation of time series can be derived.
%  Initially, in this research, handcrafted features were computed from time series, such as the mean, wavelets \cite{percival2000wavelet}, or Fourier coefficients \cite{bloomfield2004fourier}, depending on expert knowledge.
%   However, this approach is time-consuming and potentially biased by priors.
%   Observing the success of neural networks, features are automatically extracted by choosing an objective to optimize and a neural architecture tailored to the problem.
%    Taking a set of time series as input, features can be identified by solving specific supervised tasks—classification \cite{ismail2019deep}, anomaly detection \cite{blazquez2021review}, forecasting \cite{torres2021deep}— or in an unsupervised manner by deriving labels from the inputs \cite{meng2023unsupervised}.
%    \thi{Rajouter une image, et une phrase par rapport au temps}
%    Although the deep learning approach is particularly effective for supervised tasks, annotated datasets are rare, making unsupervised methods more practical.
%    These methods have proven effective in identifying different patterns in the data and providing a vector representation that segregates well into clusters \cite{meng2023unsupervised}.
%     However, they may lack precision on small and homogeneous datasets with only one typical shape, often arising from clinical experiments addressing a specific phenomenon (refer to figures).
%      While current unsupervised methods separate representations into clusters, our focus is on the inter-individual variability within one cluster.
%       Typically, we aim to perform Principal Components Analysis (PCA) \cite{bro2014principal} on individual patterns with similar shapes. The representation is dependant on the input dataset.

 %\citet{srivastava2010shape,heo2024logistic} address the representation of curves $C:\msi \to \Rset^d$, $\msi\subset \Rset$, having unitary velocity by using the Square-Root Velocity (SRV) representation.
 %However, the case of time series has not been adequately addressed, considering the special attention the time axis requires.
 %For instance, the SRV representation applies after a reparametrization in time, such that the original time evolution of the time series is not represented in the final features.
 %In this regard, this work is not an application of LDDMM to 1D curves \cite{glaunes2008large}.
 %Compared to Normalizing Flows \cite{rezende2015variational,kobyzev2020normalizing} or Continuous Normalizing Flows \cite{chen2018neural,grathwohl2019scalable,salman2018deep} for diffeomorphisms learning, the number of hyperparameters to tune is minimal, and the related optimization problem is well-posed.
 Balancing between discrete and continuous elements is a challenging task.
  In the deep learning literature \cite{chen2018neural, kidger2020neural, tzen2019neural, jia2019neural, liu2019neural, ansari2023neural}, Neural Ordinary Differential Equations (Neural ODEs) \cite{chen2018neural} learn continuous latent representations using a vector field parameterized by a neural network, serving as a continuous analogue to Residual Networks \cite{zagoruyko2016wide}.
   This approach was further enhanced by Neural Controlled Differential Equations (Neural CDEs) \cite{kidger2020neural} for handling irregular time series, functioning as continuous-time analogs of RNNs \cite{schuster1997bidirectional}.
 Extending Neural ODEs, Neural Stochastic Differential Equations (Neural SDEs) introduce regularization effects \cite{liu2019neural}, although optimization remains challenging.
  Leveraging techniques from continuous-discrete filtering theory, Ansari et al. \cite{ansari2023neural} applied succesfully Neural SDEs to irregular time series.
  Oh et al. \cite{oh2024stable} improved these results by incorporating the concept of controlled paths into the drift term, similar to how Neural CDEs outperform Neural ODEs.

>>>>

%Idée de parler d'inter individualité, spécialiser davantage
%k shape, shapelets , ye2009time

Our goal is to analyze the inter-individual variability within a time series dataset, an approach of significant interest in physiological contexts \cite{guscelli2019importance, wang2016research, bar2012studying, germain2023unsupervised}.
 Specifically, we aim to develop an unsupervised feature representation method that encodes the specificities of individual time series in comparison to a reference time series.
In physiology, examining the various "shapes" in a time series related to biological phenomena and their variations due to individual differences or pathological conditions is common.
 However, the term "shape" lacks a precise definition and is more intuitively understood as the silhouette of a pattern in a time series. In this paper, we refer to the shape of a time series as the graph of this signal.

 Although community structures with representatives can be learned in an unsupervised manner \cite{trirat2024universal, meng2023unsupervised} using contrastive loss \cite{franceschi2019unsupervised, tonekaboni2021unsupervised, meng2023unsupervised} or similarity measures \cite{asgari2023clustering, germain2023unsupervised, paparrizos2015k, ye2009time},
  the study of inter-individual variability of shapes within a cluster \cite{niennattrakul2007inaccuracies, shirato2023identifying} remains an open problem in unsupervised representation learning (URL), particularly for \textit{irregularly sampled} time series with \textit{variable lengths}.

Our work explicitly focuses on learning shape-based representation of time series.
First, we propose to view the shape of a time series not merely as its curve $\{s_t:\eqsp t\in\msi\}$, but as its graph $\msg(s)=\{(t,s(t)):\eqsp t\in \msi\}$.
   Then, building on the shape analysis literature \cite{beg2005computing,vaillant2004statistics}, we adopt the Large Deformation Diffeomorphic Metric Mapping (LDDMM) framework \cite{beg2005computing,vaillant2004statistics} to analyze these graphs.
    The core idea is to represent each element $\msg(s^j)$ of a dataset $(s^j)_{j\in[N]}$ as the transformation of a reference graph $\msg(\mathbf{s}_0)$ by a diffeomorphism $\phi_j$, i.e. $\msg(s^j) \sim \phi_j . \msg(\mathbf{s}_0)$ .
    The diffeomorphism $\phi_j$ is learned by integrating an ordinary differential equation parameterized by a Reproducing Kernel Hilbert Space (RKHS).
     The parameters $(\alpha_j)_{j\in[N]}$ encoding the diffemorphisms $(\phi_j)_{j\in[N]}$ yield the representation features of the graphs $(\msg(s^j))_{j\in[N]}$.
     Finally, these shape-encoding features can be used as inputs to any statistical or machine-learning model.

@@

     LDDMM framework is a relevant shape analysis tool to represent curves as depicted in \cite{glaunes2008large}. However, graphs of time series are a well-structured type of curve due to the inclusion of the temporal dimension that requires specific care (\Cref{fig:diffeo}).
     In a similar vein, Qiu \textit{et al} \cite{qiu2009time} proposes a method for tracking anatomical shape changes in serial images using LDDMM. They include temporal evolution, but not for the same purpose: the aim is to perform longitudinal modeling of brain images.

       Leaving the LDDMM representation, the results of \cite{srivastava2010shape,heo2024logistic} addresses the representation of curves with the Square-Root Velocity (SRV) representation.
        However, the SRV representation is applied after a reparametrization of the temporal dimension on the unit length segment.
         Consequently, the graph structure of the time series is not respected, and the original time evolution of the time series is not encoded in the final representation.
          Very recently, in a functional data analysis framework, a paper \cite{wu2024shape} (Shape-FPCA) improved by representing the original time evolution.
           Nevertheless, this method is made for \textit{continuous objects} and only applies to time series of \textit{same length}, making the estimation more sensitive to noise.
  %However, the case of time series has not been adequately addressed, considering the special attention the time axis requires.
  %For instance, the SRV representation applies after a reparametrization in time, such that the original time evolution of the time series is not represented in the final features.
% Depending on the tasks, e.g classification \cite{ismail2019deep}, anomaly detection \cite{blazquez2021review}, forecasting \cite{torres2021deep} or clustering , a meaningful vector representation of time series can be derived.
%  Initially, in this research, handcrafted features were computed from time series, such as the mean, wavelets \cite{percival2000wavelet}, or Fourier coefficients \cite{bloomfield2004fourier}, depending on expert knowledge.
%   However, this approach is time-consuming and potentially biased by priors.
%   Observing the success of neural networks, features are automatically extracted by choosing an objective to optimize and a neural architecture tailored to the problem.
%    Taking a set of time series as input, features can be identified by solving specific supervised tasks—classification \cite{ismail2019deep}, anomaly detection \cite{blazquez2021review}, forecasting \cite{torres2021deep}— or in an unsupervised manner by deriving labels from the inputs \cite{meng2023unsupervised}.
%    \thi{Rajouter une image, et une phrase par rapport au temps}
%    Although the deep learning approach is particularly effective for supervised tasks, annotated datasets are rare, making unsupervised methods more practical.
%    These methods have proven effective in identifying different patterns in the data and providing a vector representation that segregates well into clusters \cite{meng2023unsupervised}.
%     However, they may lack precision on small and homogeneous datasets with only one typical shape, often arising from clinical experiments addressing a specific phenomenon (refer to figures).
%      While current unsupervised methods separate representations into clusters, our focus is on the inter-individual variability within one cluster.
%       Typically, we aim to perform Principal Components Analysis (PCA) \cite{bro2014principal} on individual patterns with similar shapes. The representation is dependant on the input dataset.

 %\citet{srivastava2010shape,heo2024logistic} address the representation of curves $C:\msi \to \Rset^d$, $\msi\subset \Rset$, having unitary velocity by using the Square-Root Velocity (SRV) representation.
 %However, the case of time series has not been adequately addressed, considering the special attention the time axis requires.
 %For instance, the SRV representation applies after a reparametrization in time, such that the original time evolution of the time series is not represented in the final features.
 %In this regard, this work is not an application of LDDMM to 1D curves \cite{glaunes2008large}.
 %Compared to Normalizing Flows \cite{rezende2015variational,kobyzev2020normalizing} or Continuous Normalizing Flows \cite{chen2018neural,grathwohl2019scalable,salman2018deep} for diffeomorphisms learning, the number of hyperparameters to tune is minimal, and the related optimization problem is well-posed.
 Balancing between discrete and continuous elements is a challenging task.
  In the deep learning literature \cite{chen2018neural, kidger2020neural, tzen2019neural, jia2019neural, liu2019neural, ansari2023neural}, Neural Ordinary Differential Equations (Neural ODEs) \cite{chen2018neural} learn continuous latent representations using a vector field parameterized by a neural network, serving as a continuous analogue to Residual Networks \cite{zagoruyko2016wide}.
   This approach was further enhanced by Neural Controlled Differential Equations (Neural CDEs) \cite{kidger2020neural} for handling irregular time series, functioning as continuous-time analogs of RNNs \cite{schuster1997bidirectional}.
 Extending Neural ODEs, Neural Stochastic Differential Equations (Neural SDEs) introduce regularization effects \cite{liu2019neural}, although optimization remains challenging.
  Leveraging techniques from continuous-discrete filtering theory, Ansari et al. \cite{ansari2023neural} applied succesfully Neural SDEs to irregular time series.
  Oh \textit{et al.} \cite{oh2024stable} improved these results by incorporating the concept of controlled paths into the drift term, similar to how Neural CDEs outperform Neural ODEs.
  With TS-LDDMM, the representation is also driven by an ODE, but the velocity field is parameterized with kernels and optimized to have a minimal norm.
